Multiscale Modeling of the Paradoxical Roles of the Immune System in EMT-Mediated Cancer

### 2019-01-21 Paper/Paper.tex
Lots of edits
Lots of edits. Comments for Adam and I to consider. Got INFL cycling
schematic in Figure 1. Probably other stuff, too.

\section{Introduction}
Immunotherapy is slowly revolutionizing how we treat cancer.
Year-by-year, more breakthroughs in the field produce effective new drugs and therapies that are having a significant impact on patient health and survival\cite{restifo2012adoptive}.  [examples] 
There are myriad untapped avenues by which scientists can further explore the role of the immune system in regulating and combatting cancer; in addition to direct cell-cell interactions between immune and cancer cells, the means by which immune cells interact with and modulate the tumor microenvironment (TME) is an important research area. Immune cells can be understood as one constituent of the TME: responding to the environment's inflammatory signals, fighting the mutated cells present, and then restoring homeostasis in that local context\cite{de2006paradoxical}.
Understanding each step in this immune cascade in the context of the TME will be paramount in fully exploring and exploiting all options for immunotherapy.

Recent work on cancer-immune interactions has shown that the inflammatory conditions of the TME can play large and sometimes surprising roles in tumor growth.
Guo describes a model in which inflammatory state of the TME affects the cell cycles of the stem cells by increasing proliferation, decreasing apoptosis, and enhancing DNA damage that leads to mutations.
Guo uses the model to show that while inflammation does have an oncogenic effect, it can also have an onco-protective role as periods of mild inflammation are better at preventing cancer than periods of no inflammation in patients that have intermittent bouts of severe inflammation.
However, the aspect of the model that included the immune system was minimal: regardless of the inflammatory environment, any cells with sufficient pathway mutations (2 in their model) had a fixed probability of being cleared by the metabolism-immune balance response. 
In this paper, we hope to improve the model by including a more robust immune module and exploring how this axis will improve our understanding of inflammatory effects on the tumor.
In addition, we explore an extra component of Guo's stem cells: where they lie on the epithelial-mesenchymal axis.

Epithelial-to-mesenchymal transition (EMT) describes the (reversible) cell plasticity process by which epithelial cells transition into mesenchymal cells.
EMT has an important yet complicated role in tumor initiation, progression, and metastasis\cite{nieto2016emt}. % References more specific to this claim might be in this paper's references. 
As mesenchymal cells often exhibit stem-like properties, their existence has been tied to the initiation and progression of cancer\cite{nieto2016emt}. % References more specific to this claim might be in this paper's references.
Also, due to their motility, some have viewed them as a natural precursor to metastasis\cite{alderton2012metastasis}.

Herein, we explore the interconnectedness of these three key systems: the tumor, the immune system, and EMT.
In addition to the interactions already discussed, the immune system affects and is affected by EMT.
The immune system affects EMT by releasing a flurry of cytokines upon arrival in the TME \cite{terry2017new}. 
In particular, regulatory T cells (Tregs) release copious quantities of the EMT-driving transforming growth factor beta (TGF-$\beta$)\cite{terry2017new}. 
Conversely, the mesenchymal cells in the tumor downregulate certain surface markers making it more difficult for cytotoxic immune cells to form the immune synapse necessary to lyse the cancer cell.
Taking these three aspects together, all influencing the others, we can consider the simple diagram in Figure \ref{fig:3PartFig} and see just how much potential there is for exploring all these relationships.
Each of these three pieces will be briefly explored in turn for how they will influence one another.

[No subsections headings within Introduction]
\subsection{The Tumor}\label{TheTumor}
The tumor influences the immune system by its presence and by the cellular processes it engages in that result in shaping the TME.
On the one hand, the mutated DNA of tumor cells results in these cells being recognized as foreign by members of the innate immune system such as natural killer (NK) cells and dendritic cells.
It is then these cells that initiate the cascade of cytokines that result in the full activation of the body's immune system to fight the non-self cells.
On the other hand, the tumor cells actively work to shape the immune response that the body launches.
This is accomplished through its own set of cytokines, such as transforming growth factor beta (TGF-$\beta$), that work to create a tumor-supportive environment, including shifting the immune balance to be immunosuppressive by enhancing the recruitment of Treg cells. Since TGF-$\beta$ is also a driver of EMT, then through direct release of TGF-$\beta$ and recruitment of TGF-$\beta$-secreting Treg cells, cells of the tumor and the surrounding tissue are driven towards a more mesenchymal profile.

\subsection{The Immune System}\label{TheImmuneSystem}
The immune system has several competing interactions with the tumor.
While the cytotoxic immune cells, NKs and cytotoxic T cells (CTLs), seek out and lyse tumor cells, the regulatory branch of the immune system, Tregs, slows down the effector functions of these other cells.

These Tregs also drive the EMT process by releasing TGF-$\beta$ upon arriving at the tumor site.

\subsection{EMT}\label{EMT}
There are many characteristics distinctive of mesenchymal cells and still much research being done in further quantifying these and determining others.
Of interest here, the literature shows two qualities of mesenchymal cells that are of interest particularly in the context of cancer and the immune system.
The first is that mesenchymal cells are less susceptible to immune clearance\cite{terry2017new}. 
The second is that mesenchymal cells proliferate at a slower rate. Ref % Adam said he had a reference in mind for this.

The research on mesenchymal cells' immune evasiveness is quite conclusive with the research even going so far as establishing functional causes for this phenomenon.
Once cytotoxic immune cells, for example NKs and CTLs, have targeted a cell to lyse, they necessarily must make a physical connection with that cell, termed an immunological synapse\cite{terry2017new}. 
However, there is preliminary evidence that due to down-regulated surface markers on mesenchymal cells among other structural changes, they are refractory to this immunological task thus rendering them less susceptible to immune clearance\cite{terry2017new}. % 
In line with known motility of mesenchymal cells, scientists have referred to this condition as mesenchymal cells being ``slippery.''
This quality of mesenchymal cells is then quantified in our model by the parameter, MIE, for mesenchymal immune evasion.

The other facet of mesenchymal cells of importance in this research is their slower proliferation rate, and this is tied to the observation that mesenchymal cells often express a more stem-like phenotype.
Indeed, even in healthy tissue, this is readily observed and seen as a useful mechanism for homeostasis as epithelial cells can transition to mesenchymal, stem-like cells under appropriate conditions.
While many qualities are associated with being stem-like, we choose to focus on a reduced proliferation rate.
In quantifying this property, we use the model parameter `mesenchymal growth arrest', MGA.
This choice is less about associating this property with quiescence and more about creating a simple name that easily, if not roughly, captures the essence of what is being modeled.
For example, while mesenchymal proliferation reduction might be a more accurate term, thinking in terms of a negative formulation adds one additional twist to accurately interpreting results which we felt was more detrimental than the possible confusion of invoking ``quiescence.''

@@

@@

\section{Methods}
[In this section it is important to say what is old (from previous model), i.e. `following the model of XX, we....' and what is new/our addition.]

To address the questions around the relationships among cancer, the immune system, and EMT, we built a multiscale, agent-based model.
The agents in the model are the tissue cells with tumorigenic potential, that can either be mutation-free, or have mutations.
It is worth noting that we do not consider spatial dependencies between agents, i.e. we assume that cells are well-mixed as are all the relevant cytokines and other molecules acting on the system.
We model the effects of immune cell types and cytokines as continuous fields within the tumor microenvironment (TME), since here single-cell resolution is not required. 
Finally, tissue cells that can be labeled either epithelial or mesenchymal, and this labeling is fluid and depends both on the TME and the internal workings of each individual cell.

\subsection{Tissue Cells}\label{TissueCells}
The tissue cells at all times have associated with them two important quantities: a logical vector indicating pathway mutations and an EMT score.
The model has three pathways that have the potential to mutate over the course of a simulation.
There is a proliferation pathway that when mutated increases the probability of the cell proliferating.
There is an apoptosis pathway that when mutated decreases the probability of the cell undergoing apoptosis.
There is an immune evasion pathway that when mutated decreases the probability that an immune cell can effectively clear the mutated cell.
% It might be nice to have a table to summarize this information?
The EMT score will be exposited in Section \ref{EMT}.

Add differential equations? % Adam suggested this, but I'm not sure what this means.
% I think this comment applied to the next section: add the equations describing immune system

\subsection{The Immune System}\label{ImmuneSystem}
The immune system is modeled as having three active immune cell types: NKs, CTLs, and Tregs.
The NKs and CTLs act on the system by recognizing mutated cells and clearing them.
Should they clear a cell, they themselves are then deactivated and removed from the immune population.
Tregs act on the system by suppressing the effector functions of NKs and CTLs, making it less likely they can clear mutated cells.
In addition to this, Tregs release TGF-$\beta$ which further shapes the TME by pushing tissues cells more towards a mesenchymal phenotype.

\subsection{EMT}\label{EMT}
The EMT score indicates how likely the cell is begin the next cell cycle as an epithelial or a mesenchymal cell.\
The score is between $0$ and $1$ with higher values being connected to higher probability of being mesenchymal.
For each patient, there is a fixed mesenchymal threshold.
If the EMT score ends the cycle above this threshold, the cell is mesenchymal, otherwise it is epithelial for the next cycle.
For the purposes of the model, a cell in an epithelial state is considered in the base state, and one in a mesenchymal state will have some of its parameters updated.
In particular, mesenchymal cells benefit from increased immune evasion but suffer from decreased proliferation.
Note, that because the immune system can only clear mutated cells, the increased immune evasion only benefits mutated mesenchymal cells.

@@

\subsubsection{Initial conditions} Every model run starts with a fixed amount of cells, $N_0$.
However, different parameter values will lead to different steady states for the population of tissue cells.
In light of this, a warmup period of ~1000 cell cycles is used in which time no mutations are allowed to happen.
Thus, the only immune cell present is NKs.
After the warmup cycles are complete, a mutagenic event is simulated in which all the tissue cells experience a discrete, random jump in their probability to mutate.
From then on, a cell which proliferates either mutates or increases its probability of mutating later on.

\subsubsection{Tissue cell fate}
During each cell cycle, every cell randomly is assigned a cell fate from the following options:
\begin{itemize}
\item proliferation
\item apoptosis
\item immune clearance (by NKs or CTLs)
\item quiescence
\end{itemize}

@@

\paragraph{Proliferation}
There are four factors that contribute to a cell's probability of proliferation.
% NB try to avoid posessive apostrophe. I.e. replace this with
There are four factors that contribute to the probability that a cell will proliferate.
% 
The first is a base proliferation probability that all cells have, $p$.
Second, if the cell has a mutation in the proliferation pathway, then the weight for proliferation is proportionally increased by $\text{prolif}_{\text{up}}$.
Third, if the cell is mesenchymal, then the weight for proliferation is proportionally decreased by $m_Q$, which stands for mesenchymal quiescence.
Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor.
In total, the weight for proliferation is given by

$$ p_{\text{prolif}} = p(1+\delta_{\text{PM}}\text{prolif}_{\text{up}})(1-\delta_{MES}m_Q)\frac{N_0}{N_0+N} $$

@@

\paragraph{Immune Clearance}
For both NK clearance and CTL clearance, the weights are built with the same factors but have different parameter values for NK and CTLs.
First of all, the cell needs to be mutated, $\delta_M$.
Second, there is a Hill factor that captures the probability of an immune cell finding and interacting with the given tissue cell.
Third, NKs and CTLs have their own efficacy parameters which can be understood as their probability of clearing a mutated cell given that they have interacted with it.
Fourth, there is a decreasing Hill factor based on the number of Treg cells present.
Finally, there are two factors that proportionally decrease the weight of immune clearance depending on if the cell has an immune evasion mutation or if it is mesenchymal.
In total, the weight of NK clearance is given by

$$ p_{\text{NKed}} =  \delta_M \frac{\text{NK}}{N/N_{00}+\text{NK}} E_{\text{NK}} \frac{1}{1+\text{Treg}/\text{Treg}_{\text{EC50}}} $$ $$ (1-\delta_{IEM}\text{IE}_{\text{up}})(1-\delta_{MES}m_{\text{IE}})$$

@@

\paragraph{Quiescence}
The weight associated with quiescence is taken as 1 except in the case of Mesenchymal cells.
For Mesenchymal cells, the weight lost to proliferation is added to the weight of quiescence.
Hence, the weight of quiescence is given by

$$ p_Q = 1 + \delta_{\text{MES}}p(1+\delta_{\text{PM}}\text{prolif}_{\text{up}})m_Q\frac{N_0}{N_0+N} $$

The reason for adding that term is due to the understanding that mesenchymal cells are not proliferating as fast because they are instead remaining quiescent, hence naming the parameter $m_Q$.

\subsubsection{Completing the Cell Cycle}
After the cell fates are determined and the results reflected in the system, there are a few things that happen before the system moves on to a new cell cycle.
First, the NK and CTL populations are reduced by the number of mutated cells they cleared.
This represents the fact that individual immune cells have a limited amount of efficacy.

Also, all proliferating cells have a probability of undergoing a pathway mutation.
If they do, one is randomly chosen among the three and that one either mutates or remains mutated.
If they do, one pathway is randomly chosen among the three and is mutated.
If the cell does not undergo a mutation, then its probability of mutating during subsequent cell increases. [**]

Then, the EMT values for each cell is updated.
This depends on how much TGF-$\beta$ is currently in the system.
Each cell receives an amount of TGF-$\beta$ given by

@@

where $\sigma$ is a parameter in the model.
This quantity is then compared to $2-\text{EMT}_i$ where $\text{EMT}_i$ is the EMT score of the $i^{th}$ cell on a scale of $[0,1]$ with $0$ representing most epithelial and $1$ representing most mesenchymal.
If $\text{TGFB}_i>2-\text{EMT}_i$, the cell becomes more mesenchymal and will have a higher EMT score.
If $\text{TGFB}_i<2-\text{EMT}_i$, the cell becomes more epithelial and will have a lower EMT score.
The reason for the expression $2-\text{EMT}_i$ can be understood this way: 
the bounds on the EMT score guarantee $2-\text{EMT}_i\in[1,2]$.
Hence, there either needs to be a lot of TGF-$\beta$ in the system or a large amount of randomness for a cell to undergo EMT.
In addition, a cell having a higher EMT score will have a higher chance of continuing in the EMT process whereas a cell low on the EMT spectrum will have a higher chance of moving towards a more epithelial phenotype.

@@

$$ \text{TGFB} = \text{tgfb}_{\text{mut}}N_{\text{mut}} + \text{tgfb}_{\text{Treg}}\text{Treg}$$

@@

$$ \text{NK}_{i+1} = \left (\text{NK}_i-\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} \right )\exp(-d_{\text{NK}}\Delta t)+\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} $$

For CTLs and Tregs, however, they rely on antigen presenting cells (APCs) to be activated, and these APCs in turn rely on mutated cells being cleared so that they can bind antigens.
In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function.
Thus, let $\text{APC}$ represent the number of mutated cells cleared by the immune system during a given cycle.
We choose the following differential equations to govern the CTL and Treg populations:

>>>>

\section{Introduction}
Immunotherapy is slowly revolutionizing how we treat cancer.
Year-by-year, new breakthroughs in the field produce new, effective drugs and therapies that have a significant impact on patient health and survival\cite{pardoll2012blockade}\cite{restifo2012adoptive}.  % More examples? Reviews or non-reviews? Explain some of the examples?
There are myriad untapped avenues by which scientists can further explore the role of the immune system in regulating and combatting cancer; in addition to direct cell-cell interactions between immune and cancer cells, the means by which immune cells interact with and modulate the tumor microenvironment (TME) is an important research area. Immune cells can be understood as one constituent of the TME: responding to the environment's inflammatory signals, fighting the mutated cells present, and then restoring homeostasis in that local context\cite{de2006paradoxical}.
Understanding each step in this immune cascade in the context of the TME will be necessary in order to fully explore and exploit all options of immunotherapy.

Recent work on cancer-immune interactions has shown that the inflammatory conditions of the TME can play large and sometimes surprising roles in tumor growth.
Guo describes a model in which the inflammatory state of the TME affects the cell cycles of the stem cells by increasing proliferation, decreasing apoptosis, and enhancing DNA damage that leads to mutations\cite{guo2017multiscale}.
Guo uses the model to show that while inflammation does have an oncogenic effect, it can also have an onco-protective role as periods of mild inflammation are better at preventing cancer than periods of no inflammation in patients that have intermittent bouts of severe inflammation.
However, the aspect of the model that included the immune system was minimal: regardless of the inflammatory environment, any cells with sufficient pathway mutations (2 in their model) had a fixed probability of being cleared by the metabolism-immune balance response. 
In this paper, we hope to improve the model by including a more robust immune module and exploring how this axis will improve our understanding of inflammatory effects on the tumor.
In addition, we explore an extra component of Guo's stem cells: where they lie on the epithelial-mesenchymal axis.
So, in total, we will explore the joint interactions of the tumor, the immune system, and the epithelial-mesenchymal spectrum.
We will also study the effects of the TME on these interactions.

The tumor interacts with the immune system by initiating the sequence of events that results in the activation of the adaptive immune response.
It accomplishes this by presenting antigens that innate immune cells recognize and carry to lymph nodes to activate, among other things, T cells.
Also, tumor cells engage in certain cellular processes that shape the TME and thus impact the immune system.
In particular, tumor cells release transforming growth factor beta (TGF-$\beta$) that creates a tumor-supportive environment by shifting the immune balance to being immunosuppressive via enhancing activation of Treg cells.

This particular cytokine is also a driver of the epithelial-to-mesenchymal transition (EMT) and through generating TGF-$\beta$ the tumor creates a more mesenchymal phenotype within the TME.
In this way, the tumor shifts the balance of local environment along the EMT spectrum.

The immune system has several competing interactions with the tumor.
While the cytotoxic immune cells, NKs and cytotoxic T cells (CTLs), seek out and lyse tumor cells, the regulatory branch of the immune system, which will be accounted for by Tregs in our model, slows down the effector functions of these other cells.

In relation to EMT, the Tregs of the immune system drive the EMT process by releasing TGF-$\beta$ upon arriving at the tumor site\cite{terry2017new}.
In this way, just as the tumor cells push the TME towards a mesenchymal phenotype, so too do the Tregs.

Epithelial-to-mesenchymal transition (EMT) describes the (reversible) cell plasticity process by which epithelial cells transition into mesenchymal cells.
While the epithelial cells often stay bound together in layers, mesenchymal cells have gained motility and even stem-like properties\cite{nieto2016emt}.
They have thus become a subject of great interest in cancer research having been connected to tumor initiation, progression, and metastasis\cite{nieto2016emt}.
There are many other characteristics distinctive of mesenchymal cells and still much research being done in further quantifying these and determining others.
Of interest to us, the literature reveals two qualities of mesenchymal cells that are particularly relevant in the context of cancer and the immune system.
The first is that mesenchymal cells are less susceptible to immune clearance\cite{terry2017new}. 
The second is that mesenchymal cells proliferate at a slower rate. Ref % Adam said he had a reference in mind for this.

The research on mesenchymal cells' immune evasiveness is quite conclusive with the research even going so far as establishing functional causes for this phenomenon.
As a cell is being targeted by cytotoxic immune cells for clearance, a physical connection between the two cells must be created in order to precipitate the lysing event.
This immunological synapse, however, is dependent upon the surface proteins of the target cell, and in the case of mesenchymal cells there is preliminary evidence that these surface markers are down-regulated in such a way that the immune synapse is more difficult to form\cite{terry2017new}.
There are also other structural changes that take place in mesenchymal cells that could contribute to this refractory nature towards immune clearance.
Regardless, the evidence currently demonstrates a ``slipperiness'' of mesenchymal cells in regards to immune lysing that in our model we will refer to as mesenchymal immune evasion, or MIE.

The other facet of mesenchymal cells of importance in this research is their slower proliferation rate, and this is tied to the observation that mesenchymal cells often express a more stem-like phenotype.
Indeed, even in healthy tissue, this is readily observed and seen as a useful mechanism for homeostasis as epithelial cells can transition to mesenchymal, stem-like cells under appropriate conditions. Ref
While many qualities are associated with being stem-like, we choose to focus on a reduced proliferation rate. Ref
Cancer cells are well-known for achieving unregulated proliferation, so exploring a subset of cancer cells with decreased proliferation could provide intriguing new results.
In our model, we refer to this property as mesenchymal growth arrest, or MGA.

@@

%%%%%%%%%%%%%  INTRO PARAGRAPH GRAVEYARD %%%%%%%%%%%%%%%%

%They release cytokines that alter the extracellular matrix, promote angiogenesis, and create an inflammatory environment.
%This last one, in particular, results in the impairment of cytotoxic capabilities of immune cells.
%Through 
%
% its presence and by the cellular processes it engages in that result in shaping the TME.
%On the one hand, the mutated DNA of tumor cells results in these cells being recognized as foreign by members of the innate immune system such as natural killer (NK) cells and dendritic cells.
%It is then these cells that initiate the cascade of cytokines that result in the full activation of the body's immune system to fight the non-self cells.
%On the other hand, the tumor cells actively work to shape the immune response that the body launches.
%This is accomplished through its own set of cytokines, such as transforming growth factor beta (TGF-$\beta$), that work to create a tumor-supportive environment, including shifting the immune balance to be immunosuppressive by enhancing the recruitment of Treg cells. Since TGF-$\beta$ is also a driver of EMT, then through direct release of TGF-$\beta$ and recruitment of TGF-$\beta$-secreting Treg cells, cells of the tumor and the surrounding tissue are driven towards a more mesenchymal profile.

%In quantifying this property, we use the model parameter `mesenchymal growth arrest', MGA.
%This choice is less about associating this property with quiescence and more about creating a simple name that easily, if not roughly, captures the essence of what is being modeled.
%For example, while mesenchymal proliferation reduction might be a more accurate term, thinking in terms of a negative formulation adds one additional twist to accurately interpreting results which we felt was more detrimental than the possible confusion of invoking ``quiescence.''

%This plays an important yet complicated role in tumor initiation, progression, and metastasis\cite{nieto2016emt}. % References more specific to this claim might be in this paper's references. 
%As mesenchymal cells often exhibit stem-like properties, their existence in a tumor has been tied to the initiation and progression of cancer\cite{nieto2016emt}. % References more specific to this claim might be in this paper's references.
%Also, due to their motility, some have viewed them as a natural precursor to metastasis\cite{alderton2012metastasis}. 
%
%Conversely, the mesenchymal cells in the tumor downregulate certain surface markers making it more difficult for cytotoxic immune cells to form the immune synapse necessary to lyse the cancer cell\cite{terry2017new}.
%Taking these three aspects together, all influencing the others, we can consider the simple diagram in Figure \ref{fig:3PartFig} and see just how much potential there is for exploring all these relationships.
%Each of these three pieces will be briefly explored in turn for how they will influence one another.

@@

\section{Methods}
To address the questions around the relationships among cancer, the immune system, and EMT, we extended the multiscale, agent-based model of Guo to include an immune module and an EMT axis for the tissue cells.
The agents in the model are the tissue cells with tumorigenic potential.
They can either be mutation-free, or have any combination of three possible pathway mutations.
We model the immune cells as discrete variables but as one collective, in other words they are not agents themselves.
The cytokine TGF-$\beta$ is a continuous variables within the tumor microenvironment (TME).
It is worth noting that there is no spatial aspect to this model.
In doing this, we are assuming that cells are well-mixed as are all the relevant cytokines and other molecules acting on the system.
Finally, tissue cells can be labeled either epithelial or mesenchymal, and this labeling can change over time and depends both on the TME and the internal workings of each individual cell.

\subsection{Tissue Cells}\label{TissueCells}
The tissue cells at all times have associated with them two important quantities: which combination of mutations they carry and an EMT score.
The model has three pathways that have the potential to mutate over the course of a simulation.
There is a proliferation pathway that when mutated increases the probability of the cell proliferating.
There is an apoptosis pathway that when mutated decreases the probability of the cell undergoing apoptosis.
There is an immune evasion pathway that when mutated decreases the probability that an immune cell can effectively clear the mutated cell.
% It might be nice to have a table to summarize this information?
The EMT score will be exposited in Section \ref{EMT}. % make sure this is really the best decision

%Add differential equations? % Should this be done here or where I have it later?

\subsection{The Immune System}\label{ImmuneSystem}
The immune system is modeled as having three active immune cell types: NKs, CTLs, and Tregs.
The NKs and CTLs act on the system by recognizing mutated cells and clearing them.
Upon successful clearance, they themselves are then deactivated and removed from the immune population. % Should this be justified here or mentioned in the intro?
Tregs act on the system by suppressing the effector functions of NKs and CTLs, making it less likely they can clear mutated cells.
In addition to this, Tregs release TGF-$\beta$ which further shapes the TME by pushing tissues cells more towards a mesenchymal phenotype.

\subsection{EMT}\label{EMT}
Each tissue cell has an EMT score between 0 and 1.
When the score is above a fixed threshold value, the cell is labeled as mesenchymal.
Otherwise, it is below the threshold and labeled as epithelial.
For the purposes of the model, a cell in an epithelial state is considered in the base state, and one in a mesenchymal state will have some of its parameters updated.
In particular, mesenchymal cells benefit from increased immune evasion but suffer from decreased proliferation.
Note, that because the immune system can only clear mutated cells, the increased immune evasion only benefits mutated mesenchymal cells.

@@

\subsubsection{Initial conditions} Every model run starts with a fixed amount of cells, $N_0$.
However, different parameter values will lead to different steady states for the population of tissue cells.
In light of this, a warmup period of ~1000 cell cycles is used in which time no mutations are allowed to happen.
Thus, the only immune cell present is NKs.
After the warmup cycles are complete, a mutagenic event is simulated in which all the tissue cells experience a random increase in their probability to mutate.
From then on, a cell which proliferates either mutates or increases its probability of mutating later on.

\subsubsection{Tissue cell fate}
During each cell cycle, every cell randomly is assigned a cell fate from the following options:
\begin{itemize}
\item proliferation
\item apoptosis
\item immune clearance (by NKs or CTLs)
\item rest/$G_0$ % use both here?
\end{itemize}

@@

\paragraph{Proliferation}
There are four factors that contribute to the probability that a cell will proliferate.
% 
The first is a base proliferation probability that all cells have, $p$.
Second, if the cell has a mutation in the proliferation pathway, then the weight for proliferation is proportionally increased by $\text{prolif}_{\text{up}}$.
Third, if the cell is mesenchymal, then the weight for proliferation is proportionally decreased by $m_{GA}$, which stands for mesenchymal growth arrest.
Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor.
In total, the weight for proliferation is given by

$$ p_{\text{prolif}} = p(1+\delta_{\text{PM}}\text{prolif}_{\text{up}})(1-\delta_{MES}m_{GA})\frac{N_0}{N_0+N} $$

@@

\paragraph{Immune Clearance}
For both NK clearance and CTL clearance, the weights are built with the same factors but have different parameter values for NK and CTLs.
First of all, the cell needs to be mutated, $\delta_M$.
Second, there is a Hill factor that captures the probability of an immune cell finding and interacting with the given tissue cell.
Third, NKs and CTLs have their own efficacy parameters, $E_{\text{NK}}$ and $E_{\text{CTL}}$, which can be understood as the weight for immune clearance given an immune cell has found the mutated cell.
Fourth, there is a decreasing Hill factor based on the number of Treg cells present.
Finally, there are two factors that proportionally decrease the weight of immune clearance depending on if the cell has an immune evasion mutation or if it is mesenchymal.
In total, the weight of NK clearance is given by

$$ p_{\text{NK}} =  \delta_M \frac{\text{NK}}{N/N_{1}+\text{NK}} E_{\text{NK}} \frac{1}{1+\text{Treg}/\text{Treg}_{\text{EC50}}} \cdot $$
$$ \qquad(1-\delta_{IEM}\text{IE}_{\text{up}})(1-\delta_{MES}m_{\text{IE}})$$
% Should I worry about the order in which these appear? Rearranging would make the formula look better, but the order I have introduced them makes the most sense.
% Also, any thoughts on NK, CTL, Treg not being italicized? While other capital letters are?

@@

\paragraph{Rest/$G_0$} %use both for this section heading?
The weight associated with quiescence is taken as 1 except in the case of Mesenchymal cells.
For Mesenchymal cells, the weight lost to proliferation is added to the weight of quiescence.
Hence, the weight of quiescence is given by

$$ p_{G0} = 1 + \delta_{\text{MES}}p(1+\delta_{\text{PM}}\text{prolif}_{\text{up}})m_{GA}\frac{N_0}{N_0+N} $$ %p_{G_0},p_{rest},p_{\text{rest}},p_{GA}???

The reason for adding that term is due to the understanding that mesenchymal cells are not proliferating as fast because they are instead remaining in the $G_0$ phase. % is this the right way to word this?

\subsubsection{Completing the Cell Cycle}
After the cell fates are determined and the results reflected in the system, there are a few things that happen before the system moves on to a new cell cycle.
First, the NK and CTL populations are reduced by the number of mutated cells they cleared.
This represents the fact that individual immune cells lose efficacy as they carry out their effector functions. % Similar to previous comment: should I cite reference for this here or put this up in the intro?

Also, all proliferating cells have a probability of undergoing a pathway mutation.
If they do, one is randomly chosen among the three and that one either mutates or remains mutated.
If the cell does not undergo a mutation, then its probability of mutating during subsequent cell increases. % Updated this after Adam had question. Does this make sense as is?

Then, the EMT values for each cell is updated.
This depends on the cells current EMT score and how much TGF-$\beta$ is currently in the system.
Each cell receives an amount of TGF-$\beta$ given by

@@

where $\sigma$ is a parameter in the model.
If this quantity is large enough, relative to the EMT score of the cell, then the cell will undergo EMT and its EMT score will increase.
If not, then the cell undergoes the reverse process, MET, and its EMT score will decrease.
Each cell then is relabeled as either epithelial or mesenchymal depending on its new EMT score and whether it is below or above the mesenchymal threshold.

%%%%%%%%%	I think this next section is too much detail that just confuses the reader. I don't think I need to explain my implementation, just the process. Thoughts?	%%%%%%%%%%%%%%%%%%%%
This quantity is then compared to $2-\text{EMT}_i$ where $\text{EMT}_i$ is the EMT score of the $i^{th}$ cell on a scale of $[0,1]$ with $0$ representing most epithelial and $1$ representing most mesenchymal.
If $\text{TGFB}_i>2-\text{EMT}_i$, the cell becomes more mesenchymal and will have a higher EMT score.
If $\text{TGFB}_i<2-\text{EMT}_i$, the cell becomes more epithelial and will have a lower EMT score.
The reason for the expression $2-\text{EMT}_i$ can be understood this way: 
the bounds on the EMT score guarantee $2-\text{EMT}_i\in[1,2]$.
Hence, there either needs to be a lot of TGF-$\beta$ in the system or a large amount of randomness for a cell to undergo EMT.
In addition, a cell having a higher EMT score will have a higher chance of continuing in the EMT process whereas a cell low on the EMT spectrum will have a higher chance of moving towards a more epithelial phenotype.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

$$ \text{TGFB}_k = \text{tgfb}_{\text{mut}}M_k + \text{tgfb}_{\text{Treg}}\text{Treg}_k$$ % clear that M is number of mutants, right?

@@

%%% I like the first of these two options. You set aside the amount you know you will gain (sigma/d) and then let the rest decay exponentially
$$ \text{NK}_{k+1} = \left (\text{NK}_k-\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} \right )\exp(-d_{\text{NK}}\Delta t)+\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} $$

%%% For this one, the initial decays exponentially and then you add some fraction of sigma/d depending on how big your time step is
$$ \text{NK}_{k+1} = \text{NK}_k\exp(-d_{\text{NK}}\Delta t)+\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} \left(1-\exp(-d_{\text{NK}}\Delta t)\right)$$
%%% The first one is better though in my opinion.

% For all 3 discretizations of immune cell dynamics, I could create a quantity R_* (for recruitment? maybe better letters exist) so the formula would be 
% *_{k+1} = (*_k-R_*)\exp(-d_*\Delta t)+R_*,\quad R_*=\frac{\sigma_*}{d_*}
%% Option 1b from above:
$$\text{NK}_{k+1} = (\text{NK}_k-R_\text{NK})\exp(-d_\text{NK}\Delta t)+R_\text{NK},\quad R_\text{NK}=\frac{\sigma_\text{NK}}{d_\text{NK}}$$

%%%% Which of these two paragraphs is better? Currently the intro does not mention APCs and I don't think I need to go into that, especially when APC_k = \tilde{M}_k
For CTLs and Tregs, they rely on mutant cells being cleared before they can be activated.
Let $\tilde{M}_k$ represent the number of mutant cells cleared by the immune system during cell cycle $k$.
In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function.
We choose the following differential equations to govern the CTL and Treg populations:

\begin{align*}
\text{CTL}' & = \sigma_{\text{CTL}}\tilde{M} - d_{\text{CTL}}\text{CTL} \\
\text{Treg}' & = \sigma_{\text{Treg}}\tilde{M} \frac{\text{TGFB}}{1+\text{TGFB}/\text{EC50}}- d_{\text{Treg}}\text{Treg}
\end{align*}

\begin{align*}
\text{CTL}_{k+1} & = & \left (\text{CTL}_k-\sigma_{\text{CTL}}\tilde{M}_k/d_{\text{CTL}}\right )\exp(- d_{\text{CTL}}\Delta t) + \sigma_{\text{CTL}}\tilde{M}_k/d_{\text{CTL}}\\
\text{Treg}_{k+1} & = & \left (\text{Treg}_k-\frac{\sigma_{\text{Treg}}\tilde{M}_k}{d_{\text{Treg}}} \frac{\text{TGFB}}{1+\text{TGFB}/\text{EC50}}\right )\exp(-d_{\text{Treg}}\Delta t)+\\
& & \frac{\sigma_{\text{Treg}}\tilde{M}_k}{d_{\text{Treg}}} \frac{\text{TGFB}}{1+\text{TGFB}/\text{EC50}}
\end{align*}

%% Or using R_*
\begin{align*}
\text{CTL}_{k+1} & =  \left (\text{CTL}_k-\tilde{M}_kR_{\text{CTL}}\right )\exp(- d_{\text{CTL}}\Delta t) + \tilde{M}_kR_{\text{CTL}}\\
\text{Treg}_{k+1} & =  \left (\text{Treg}_k-\tilde{M}_kR_{\text{Treg}}\right )\exp(-d_{\text{Treg}}\Delta t)+\tilde{M}_kR_{\text{Treg}}
\end{align*}
where
\begin{align*}
R_{\text{CTL}} &= \sigma_{\text{CTL}}/d_{\text{CTL}}\\
R_{\text{Treg}} &=\frac{\sigma_{\text{Treg}}}{d_{\text{Treg}}} \frac{\text{TGFB}}{1+\text{TGFB}/\text{EC50}}
\end{align*}
%% End of option 1

%% Beginning of option 2
For CTLs and Tregs, however, they rely on antigen presenting cells (APCs) to be activated, and these APCs in turn rely on mutated cells being cleared so that they can bind antigens.
In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function.
Thus, let $\text{APC}$ represent the number of mutated cells cleared by the immune system during a given cycle.
We choose the following differential equations to govern the CTL and Treg populations:
